### 2015-11-29 paper/1-introduction.tex

\subsection{Parallel computing}
%TODO parallel computing is necessary with more and more complex systems, blablabla (usual explanation)
Parallel execution of a PDEVS simulation can lower runtime and increase the bound on the state space, thereby enabling simulation of more complex systems in the same time-frame.
While the shared memory parallelism offered by most modern hardware does not increase the state space bounds, it can reduce the runtime and offers more direct communication and control between entities involved in synchronization compared to distributed simulation. \\

\subsection{Motivation}
%TODO give motivation (adevs is good, but only conservative; exploit new parallelism features of C++11 too)
Adevs \cite{adevs} offers a very fast conservative synchronized shared memory DEVS simulator, but no optimistic synchronized variant. 
The latter can be significantly faster, especially in simulations where the runtime behaviour of the simulation is hard to predict. \\
The matured parallelism features of C++11 were used in this project with the dual aim of writing standard-compliant (and thus portable) code without losing access to powerful low-level threading primitives. % may need another pass on this one :(

>>>>

\subsection{Parallel computing}
%TODO parallel computing is necessary with more and more complex systems, blablabla (usual explanation)
Parallel execution of a PDEVS simulation can lower overall runtime and increase the bound on the state space, thereby enabling simulation of more complex systems in the same time-frame.
While the shared memory parallelism offered by most modern hardware does not increase the state space bounds, it can reduce the runtime and offers more direct communication and control between entities involved in synchronization compared to distributed simulation. \\

\subsection{Motivation}
%TODO give motivation (adevs is good, but only conservative; exploit new parallelism features of C++11 too)
Adevs \cite{adevs} offers a very fast conservative synchronized shared memory DEVS simulator, but no optimistic synchronized variant. 
The latter can be significantly faster, especially in simulations where the runtime behaviour of the simulation is hard to predict. \\
The matured parallelism features of C++11 were used in this project with the dual aim of writing standard-compliant (and thus portable) code and without losing access to powerful low-level threading primitives. % may need another pass on this one :(

### 2015-11-29 paper/2-background.tex

\subsection{Conservative Synchronization}
%TODO explain conservative synchronization, and how it uses locking to wait for continuing the simulation
Conservative synchronization is defined by the invariant that no model will advance in time before it has received all input from any influencing model. \\
This requires the concepts of earliest output time (eot) and earliest input time (eit), which define the timespan within which a model can safely advance. \\ The eit of any model is the minimum of all eot values of (directly) influencing models. A model can simulate up to (but not including) eit, then waits until that value is increased. An important disadvantage here is that the influenced-by relation is always defined at model(link) creation, not at runtime. A model that can influence another, but never does, can severely slow down the protocol.\\ 
Deadlock between models that influence each other and end up waiting on each other can be broken/avoided by a variety of means. In this simulator, the CMB \cite{Chandy:1981:ADS:358598.358613} null-message protocol is used. \\
In our implementation null-time is the timestamp a model is guaranteed to have passed in simulation. More precisely, a null message of time t is a guarantee that any output with timestamp t-$\epsilon$ is already sent.\\
In general, the eot/eit/nulltime of a kernel is the minimum of each of those values for all models in the kernel.\\
Conservative synchronization relies explicitly on information provided by the model creator in the form of a lookahead, a relative timespan during which the model is insensitive to outside events. This lookahead can be non-trivial to calculate. In general, a simulation writer will not be able to predict the exact lookahead of models involved in an experiment without having run the experiment.\\ Conservative kernels can operate if there exists a cyclic dependency between them, but at a quite severe performance penalty, as seen in section 4.
\subsection{Optimistic Synchronization}
%TODO explain optimistic synchronization, and how it uses rollbacks to undo causality errors
Optimistic synchronization allows causality errors to occur but recovers from those errors using a roll-back mechanism, the most common of which is Timewarp \cite{Jefferson:1985:VT:3916.3988}.
Whenever a kernel receives an event with a timestamp in the kernel's past, the state of the kernel (and all models) is reverted to that time. The possible gain in runtime is offset by the increase in memory required to keep saved states and (sent) messages.
In contrast to conservative, optimistic does not rely on any domain specific information. It is only sensitive to runtime use of connections, not the probability that they might occur. %reword connections ?
\\
If the (runtime) dependency graph contains a cycle, optimistic can suffer a series of cascading reverts. Without domain specific information, the kernel assumes that any event will influence at least one model, but this can lead to an infinite loop of reverts in the worst case.\\
This effect can be lessened by lazy cancellation and/or lazy re-evaluation \cite{FujimotoBook}.
\subsection{Global Virtual Time}
To avoid exhausting memory in state/event saving, optimistic synchronization relies on the concept of global virtual time\cite{Jefferson:1985:VT:3916.3988}. In optimistic simulations, GVT is defined as the lowest timestamp of any unprocessed event. \\ Intuitively this is the simulation timepoint that is certain to be preserved, corresponding exactly with the simulation up to that time in a non-parallel implementation.\\
In conservative the minimum simulation time of all kernels is the GVT, or in terms of null messages: the least timestamp of any null message in transit.
The GVT calculation is vital to safely commit unrecoverable transactions such as IO (e.g. tracing), releasing memory, ... .

>>>>

\subsection{Conservative Synchronization}
%TODO explain conservative synchronization, and how it uses locking to wait for continuing the simulation
Conservative synchronization is defined by the invariant that no model will advance in time before it has received all input from any influencing model. \\
This requires the concepts of earliest output time (eot) and earliest input time (eit), which define the timespan within which a model can safely advance. \\ The eit of any model is the minimum of all eot values of (directly) influencing models. A model can simulate up to (but not including) eit, then waits until that value is increased. An important disadvantage here is that the influenced-by relation is always defined at model(link) creation, not at runtime. A model that can influence another, but never does, can severely slow down the protocol.\\ 
Deadlock between models that influence each other and end up waiting on each other can be broken/avoided by a variety of means. In this simulator, the CMB \cite{Chandy:1981:ADS:358598.358613} null-message protocol is used. \\
In our implementation null-time is the timestamp a model is guaranteed to have passed in simulation. More precisely, a null message of time t is a guarantee that any output with timestamp t-$\epsilon$ is already sent.\\
In general, the eot/eit/nulltime of a kernel is the minimum of each of those values for all models in the kernel.\\
Conservative synchronization relies explicitly on information provided by the model creator in the form of a lookahead, a relative timespan during which the model is insensitive to outside events. This lookahead can be non-trivial to calculate. In general, a simulation writer will not be able to predict the exact lookahead of models involved in an experiment without having run the experiment.\\ Conservative kernels can operate if there exists a cyclic dependency between them, but at a quite severe performance penalty, as seen in section 4.
\subsection{Optimistic Synchronization}
%TODO explain optimistic synchronization, and how it uses rollbacks to undo causality errors
Optimistic synchronization allows causality errors to occur but recovers from those errors using a roll-back mechanism, the most common of which is Timewarp \cite{Jefferson:1985:VT:3916.3988}.
Whenever a kernel receives an event with a timestamp in the kernel's past, the state of the kernel (and all models) is reverted to that time. The possible gain in runtime is offset by the increase in memory required to keep saved states and (sent) messages.
In contrast to conservative, optimistic does not rely on any domain specific information. It is only sensitive to runtime use of connections, not the probability that they might occur. %reword connections ?
\\
If the (runtime) dependency graph contains a cycle, optimistic can suffer a series of cascading reverts. Without domain specific information, the kernel assumes that any event will influence at least one model, but this can lead to an infinite loop of reverts in the worst case.\\
This effect can be lessened by lazy cancellation and/or lazy re-evaluation \cite{FujimotoBook}.
\subsection{Global Virtual Time}
To avoid exhausting memory in state/event saving, optimistic synchronization relies on the concept of global virtual time\cite{Jefferson:1985:VT:3916.3988}. In optimistic simulations, GVT is defined as the lowest timestamp of any unprocessed event. \\ Intuitively this is the simulation timepoint that is certain to be preserved, corresponding exactly with the simulation up to that time in a non-parallel implementation.\\
In conservative, the minimum simulation time of all kernels is the GVT, or in terms of null messages: the least timestamp of any null message in transit.
The GVT calculation is vital to safely commit unrecoverable transactions such as IO (e.g. tracing), releasing memory, ... .

### 2015-11-29 paper/3-features.tex

%TODO explain that we are briefly going over the main features
\subsection{Based on PythonPDEVS}
%TODO say that it is based on PythonPDEVS
The simulator is based on PythonPDEVS, and provides the following features: 
\begin{enumerate}
	\item Direct Connection
	\item Dynamic Structured DEVS
	\item Termination function. If specified, a termination function is applied every simulation round to each model to test whether the simulation can terminate. Only available in single-threaded simulation.
	\item State/Message can have any payload type. Different message types can be used together within the same simulation.
	\item Tracing An asynchronous, thread safe and versatile tracing mechanism allows exact verification of the simulation.
	\item Optimistic and Conservative synchronization of PDEVS.
\end{enumerate}
The implementation tries to adhere to the C++ principle that you don't pay performance-wise for what you don't use. For this reason, the support for a termination function for the multi-threaded kernel was abandoned, as it is non-trivial to implement and had a non-negligible impact on the runtime, even when not in use.\\
The tracing is not comparable with adevs's listener interface. To be usable in optimistic simulation, the tracing of the simulation has to be reversible and only be committed at GVT points. Furthermore, the framework itself has to be thread safe and deterministic so that a simulation will always produce the exact same output.
%TODO Decide whether we should list the functionality that is _not_ implemented
The following features from PyPDEVS are not present
\begin{enumerate}
	\item Activity tracking and relocation
	\item Serialization
	\item Interactive control. 
	\item Distributed simulation
\end{enumerate}
Serialization in this context is the ability to save/load a complete simulation to disk, not the state saving mechanisms required for TimeWarp.\\ State saving has no impact in a single threaded or conservative kernel.\\
Model allocation is done by a derivable allocator object which the user can implement to arrange a more ideal (domain-specific) allocation. If this is omitted, a default (non-activity-aware) allocation stripes the models over the simulation kernels.\\
Debugging tools such as a logger and a graph visualizer are included which can track activity with respect to allocation for later study, but not online/dynamic as is possible in PyPDEVS.\\
While PyPDEVS can be controlled from within a Python script and adevs has a Java interface, our implementation does not have any bindings to other languages.

%TODO ... and thus has most features from PythonPDEVS, like Dynamic Structure, termination condition, direct connection, ...
%TODO (size: take as much size as you want, depending on how much is shared with PythonPDEVS)
%TODO (maybe: note that models are not compatible)
\subsection{Different Synchronization protocols}
%TODO explain rationale: why different synchronization protocols
% weak scenarios for C, O and both ?
\subsubsection{Conservative}
% This needs better explaining, perhaps a central graph to reference to ?
A conservative kernel will determine which kernels it is influenced by. This information is constructed from the incoming connections on all hosted models. The process is only 1 link deep, since an influenced kernel will in turn be blocked by others deeper in the graph.\\
A model should provide a lookahead function which returns, relative from the current time, the timespan during which the model cannot change state due to an external event. This information is collected for all models hosted on the kernel, and the minimum is set as the lookahead of that kernel. \\
The kernel will calculate its earliest output time and write this value in shared memory. The eit of the kernel is then set as the minimal eot of all influencing kernels. \\
For garbage collection (of sent messages) the LBTS/GVT is calculated as $\min_{\forall i \in \textup{influencors}}( \textup{nulltime}[i]\ )  - \epsilon $. %This is one of the operations that can benefit from using a relaxed memory ordering. % Using it for read atm, check before finalizing and do we need to mention this ?.

@@

\subsection{Performance Improvements}
%TODO briefly discuss some of the additional changes you implemented
%TODO such as memory pools, memory allocators, profiling-guided-optimizations, ...
Continuous profiling of the kernels in several benchmarks highlighted the following key bottlenecks: 
\subsubsection{Heap}
A kernel never sends a complete object to another kernel, only a pointer to the object. This avoids a possibly expensive copy of the payload, but at the cost of allocation overhead.\\
This cost becomes prohibitively expensive in highly connected models, so to reduce that overhead we use thread\_local memory pools for states and messages, and optionally replace the system malloc with calls to tcmalloc\cite{tcmalloc}. In this way allocating threads do not block each other, and in a single threaded kernel we can leverage arena-style pools if desired.\\
This changed the ownership semantics of several objects in a non-trivial way, since the thread that creates an object has to destroy it (iff it can prove it is no longer used). Experiments with synchronized pools proved slower than malloc/free. \\
Initially the kernels used strings as identifiers, as is done in PyPDEVS, profiling quickly indicated this to be a performance bottleneck. C++ strings are heap allocated variable sized objects with an atomic reference count. Access of that reference count across threads is expensive, as are the calls to malloc/free the string implementation makes to create/destroy new object, or copy existing.\\
Strings are more intuitive to work with from a user's standpoint, so as a compromise the user can reference models/ports by string name (usually when constructing the model). Once simulation starts all objects use integral identifiers for performance. This also increased usage of the constexpr feature of C++11 in, amongst others, timestamps and message headers.
\\
\subsubsection{Smart pointers}
While an important C++11 feature in general, our initial usage of smart pointers for some types of objects was misplaced. Used across threads the reference counting becomes prohibitively expensive, and the (de)allocating caused significant contention between threads. Models are still held by a smart pointer, as is a kernel, but a message is a raw pointer to compacted memory. 
\subsubsection{Locking}
Locking between kernels uses mostly atomic operations, where we can occasionally leverage memory orderings to only pay for synchronization when we need it. Messages are exchanged via a shared set of queues each with a dedicated lock.\\
On a higher level, we avoid the sending of synchronization messages entirely by writing the timestamp directly into shared memory.\\ Sending of antimessages is fairly cheap in our implementation, since only the modified pointer to the original message is sent to the receiving kernel.
\subsubsection{Schedulers}
PyPDEVS has a wide range of schedulers for the user to choose from, with performance of each depending on the simulation type. Profiling showed in our case that, for a c++ implementation, the heap implementation used in adevs was faster than any of the schedulers we had tested before. Unlike most node based heaps, this scheduler uses a fixed size array where a heap is rebuilt or modified in place depending on the amount of items to update. Items are only updated, never removed.

>>>>

%TODO explain that we are briefly going over the main features
\subsection{Based on PythonPDEVS}
%TODO say that it is based on PythonPDEVS
The simulator is based on PythonPDEVS, and provides the following features: 
\begin{enumerate}
	\item Direct Connection
	\item Dynamic Structured DEVS
	\item Termination function. If specified, a termination function is applied every simulation round to each model to test whether the simulation can terminate. Only available in single-threaded simulation.
	\item State/Message can have any payload type. Different message types can be used together within the same simulation.
	\item Tracing An asynchronous, thread safe and versatile tracing mechanism allows exact verification of the simulation.
	\item Optimistic and Conservative synchronization of PDEVS.
\end{enumerate}
The implementation tries to adhere to the C++ principle that you don't pay performance-wise for what you don't use. For this reason, the support for a termination function for the multi-threaded kernel was abandoned, as it is non-trivial to implement and had a non-negligible impact on the runtime, even when not in use.\\
The tracing is not comparable to adevs's listener interface. To be usable in optimistic simulation, the tracing of the simulation has to be reversible and only be committed at GVT points. Furthermore, the framework itself has to be thread safe and deterministic so that a simulation will always produce the exact same output.
%TODO Decide whether we should list the functionality that is _not_ implemented
The following features from PyPDEVS are not present
\begin{enumerate}
	\item Activity tracking and relocation
	\item Serialization
	\item Interactive control. 
	\item Distributed simulation
\end{enumerate}
Serialization in this context is the ability to save/load a complete simulation to disk, not the state saving mechanisms required for TimeWarp.\\ State saving has no impact in a single threaded or conservative kernel.\\
Model allocation is done by a derivable allocator object which the user can implement to arrange a more ideal (domain-specific) allocation. If this is omitted, a default (non-activity-aware) allocation stripes the models over the simulation kernels.\\
Debugging tools such as a logger and a graph visualizer are included which can track activity with respect to allocation for later study, but not online/dynamic as is possible in PyPDEVS.\\
While PyPDEVS can be controlled from within a Python script and adevs has a Java interface, our implementation does not have any bindings to other languages.

%TODO ... and thus has most features from PythonPDEVS, like Dynamic Structure, termination condition, direct connection, ...
%TODO (size: take as much size as you want, depending on how much is shared with PythonPDEVS)
%TODO (maybe: note that models are not compatible)
\subsection{Different Synchronization protocols}
%TODO explain rationale: why different synchronization protocols
% weak scenarios for C, O and both ?
\subsubsection{Conservative}
% This needs better explaining, perhaps a central graph to reference to ?
A conservative kernel will determine which kernels it is influenced by. This information is constructed from the incoming connections on all hosted models. The process is only 1 link deep, since an influenced kernel will in turn be blocked by others deeper in the graph.\\
A model should provide a lookahead function which returns, relative from the current time, the timespan during which the model cannot change state due to an external event. This information is collected for all models hosted on the kernel, and the minimum is set as the lookahead of that kernel. \\
The kernel will calculate its earliest output time and write this value in shared memory. The eit of the kernel is then set as the minimal eot of all influencing kernels. \\
For garbage collection (of sent messages) the LBTS/GVT is calculated as $\min_{\forall i \in \textup{influencors}}( \textup{nulltime}[i])  - \epsilon $. %This is one of the operations that can benefit from using a relaxed memory ordering. % Using it for read atm, check before finalizing and do we need to mention this ?.

@@

\subsection{Performance Improvements}
%TODO briefly discuss some of the additional changes you implemented
%TODO such as memory pools, memory allocators, profiling-guided-optimizations, ...
Continuous profiling of the kernels in several benchmarks highlighted the following key bottlenecks: 
\subsubsection{Heap}
A kernel never sends a complete object to another kernel, only a pointer to the object. This avoids a possibly expensive copy of the payload, but at the cost of allocation overhead.\\
This cost becomes prohibitively expensive in highly connected models, so to reduce that overhead we use thread\_local memory pools for states and messages, and optionally replace the system malloc with calls to tcmalloc\cite{tcmalloc}. In this way allocating threads do not block each other, and in a single threaded kernel we can leverage arena-style pools if desired.\\
This changed the ownership semantics of several objects in a non-trivial way, since the thread that creates an object has to destroy it (if it can prove it is no longer used). Experiments with synchronized pools proved slower than malloc/free. \\
Initially the kernels used strings as identifiers, as is done in PyPDEVS, profiling quickly indicated this to be a performance bottleneck. C++ strings are heap allocated variable sized objects with an atomic reference count. Access of that reference count across threads is expensive, as are the calls to malloc/free the string implementation makes to create/destroy new object, or copy existing.\\
Strings are more intuitive to work with from a user's standpoint. So as a compromise the user can reference models/ports by string name (usually when constructing the model). Once simulation starts all objects use integral identifiers for performance. This also increased usage of the constexpr feature of C++11 in, amongst others, timestamps and message headers.
\\
\subsubsection{Raw pointers}
While an important C++11 feature in general, our initial usage of smart pointers for some types of objects was misplaced. Used across threads the reference counting becomes prohibitively expensive, and the (de)allocating caused significant contention between threads. Models are still held by a smart pointer, as is a kernel, but a message is a raw pointer to compacted memory. 
\subsubsection{Locking}
Locking between kernels uses mostly atomic operations, where we can occasionally leverage memory orderings to only pay for synchronization when we need it. Messages are exchanged via a shared set of queues each with a dedicated lock.\\
On a higher level, we avoid the sending of synchronization messages entirely by writing the timestamp directly into shared memory.\\ Sending of antimessages is fairly cheap in our implementation, since only the modified pointer to the original message is sent to the receiving kernel.
\subsubsection{Schedulers}
PyPDEVS has a wide range of schedulers for the user to choose from, with performance of each depending on the simulation type. Profiling showed in our case that, for a c++ implementation, the heap implementation used in adevs was faster than any of the schedulers we had tested before. Unlike most node based heaps, this scheduler uses a fixed size array where a heap is rebuilt or modified in place depending on the amount of items to update. Items are only updated, never removed.

### 2015-11-29 paper/4-performance.tex

\subsection{Parallel Simulation}
%TODO explain that we will now compare different synchronization protocols
%TODO compare different synchronization protocols with adevs' conservative, showing that we are faster (or at least, not that much slower) than adevs
% split: dstone/pqueue are dag's
\subsubsection{Devstone}
% Initial.
The flattened models are allocated to kernels by giving each kernel a distinct section of the chain, resulting in a low ratio of inter-kernel to intra-kernel messages. For optimistic, this can cause more reverts since the kernels will start to drift faster as the modelcount increases. Furthermore, optimistic is quite sensitive to an increase in kernels, since the delay before a revert propagates increases. Of note as well is the warm-up time this benchmark requires, for $n=d\times w$ models, it takes $\textup{timeadvance}()*n$ transitions to activate the last model in the chain. For parallel this can be reduced to $\frac{ta()*n}{kernels}$ before the last kernel becomes active.
\subsubsection{PHold}
%Phold
In Phold \cite{PHOLD} Allocation is specified in the benchmark itself, each kernel manages a single node with a constant set of sub-nodes. The parameter R determines the percentage of remote destination models.\\
The dynamic dependency graph is a very sparse version of the static dependency graph, penalizing conservative. Lookahead is $\epsilon$, so conservative spends most of its time crawling in steps of $\epsilon$. Since the dependency graph between kernels is a complete graph, this is not a simulation that scales in our implementation. For N kernels, each kernel has to query the null-time of N-1 kernels, resulting in O($N^2$) polling behaviour. This benchmark highlights therefore the price that we pay for sharing those values instead of sending an actual null message. In a non-cyclic simulation with a non-trivial lookahead (e.g. devstone), that choice however does pay off.\\
Optimistic suffers little from the above problems, however due to the high interconnectivity a cascading revert is still possible. More seriously, a revert is very expensive in PHold due to our usage of C++11's random nr generators. The cost of a revert is dominated by the recalculation of destination models, not in allocating/deallocating states/messages. Again, this could be significantly reduced using lazy-cancellation. Once a revert happens the drift between the kernels increases fast, increasing the likelihood of more reverts.

>>>>

\subsection{Parallel Simulation}
%TODO explain that we will now compare different synchronization protocols
%TODO compare different synchronization protocols with adevs' conservative, showing that we are faster (or at least, not that much slower) than adevs
% split: dstone/pqueue are dag's
\subsubsection{Devstone}
% Initial.
The flattened models are allocated to kernels by giving each kernel a distinct section of the chain, resulting in a low ratio of inter-kernel to intra-kernel messages. For optimistic, this can cause more reverts since the kernels will start to drift faster as the modelcount increases. Furthermore, optimistic is quite sensitive to an increase in kernels, since the delay before a revert propagates increases. Of note as well is the warm-up time this benchmark requires, for $n=d\times w$ models, it takes $\textup{timeadvance}()*n$ transitions to activate the last model in the chain. For parallel this can be reduced to $\frac{ta()*n}{kernels}$ before the last kernel becomes active.
\subsubsection{PHold}
%Phold
In Phold \cite{PHOLD} allocation is specified in the benchmark itself, each kernel manages a single node with a constant set of sub-nodes. The parameter R determines the percentage of remote destination models.\\
The dynamic dependency graph is a very sparse version of the static dependency graph, penalizing conservative. Lookahead is $\epsilon$, so conservative spends most of its time crawling in steps of $\epsilon$. Since the dependency graph between kernels is a complete graph, this is not a simulation that scales in our implementation. For N kernels, each kernel has to query the null-time of N-1 kernels, resulting in O($N^2$) polling behaviour. This benchmark therefore highlights the price that we pay for sharing those values instead of sending an actual null message. In a non-cyclic simulation with a non-trivial lookahead (e.g. devstone), that choice however does pay off.\\
Optimistic suffers little from the above problems, however due to the high interconnectivity a cascading revert is still possible. More seriously, a revert is very expensive in PHold due to our usage of C++11's random nr generators. The cost of a revert is dominated by the recalculation of destination models, not in allocating/deallocating states/messages. Again, this could be significantly reduced using lazy-cancellation. Once a revert happens the drift between the kernels increases fast, increasing the likelihood of more reverts.
